---
title: "Playing with first merged crdt"
format: html
editor: visual
---

# Setting up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
library(MASS)
mergecov<-read.csv(here("data","first_merged_CRDT.csv"))
mergecov$Date<-ymd(mergecov$Date)
mergecov$Race<-as.factor(mergecov$Race)
mergecov$State<-as.factor(mergecov$State)
mergecov$Region<-as.factor(mergecov$Region)
mergecov$Gov_Control<-as.factor(mergecov$Gov_Control)
mergecov<-mergecov|>dplyr::select(c(2,3,4,5,6,7,8,9,10,13))
glimpse(mergecov)
```

## Definitions of Variables

Data set created by Covidtracking.com

-   Date\~

-   State\~

-   Race: race of people referred to by entry \~ "Total" refers to all groups

-   Cases: confirmed plus probable cases \~ person who positive test for covid, had a positive antigen test, had clinical and epidemiological evidence of covid, or had covid listed on death certificate

-   Deaths: deaths of people confirmed with or probable to have had covid \~ typically having covid on death certificate

-   Hosp: people who have ever been hospitalized with covid

-   Tests: estimate of PCR tests given

-   Region: states were divided into 7 geographical regions

-   Population: population of state provided by the "tidycensus" package

# Guiding Questions

## Overarching

-   How is any individual affected differently by a pandemic disease, like COVID-19, based on their personal identities, mainly race?

## General

-   What is the sociological intersection like between COVID deaths, state government jurisdiction, and race? How is the number of deaths affected buy these factors?

-   Does geographic region affect the number of cases? Perhaps due to environmental implications

-   Are people of certain races accessing medical care (tests and hospitals) differently? Could this answer be related to governmental jurisdictions in their state?

# Stats

## Getting the Data We Need

First, I'll just grab the totals

```{r}
covtotals<-mergecov|>filter(Race=="Total",State!="Territory")#Territory data is too sparse to consider
```

Next, I'll grab the most recent date

```{r}
covtotalslast<-covtotals|>filter(Date=="2021-03-07")
```

Now let's look at the distribution of death counts. To make it easier to read, let's look specifically at the death rate per 10,000 people

```{r}
ggplot(covtotalslast,aes(10000*Deaths/Population))+
  geom_histogram()
```

The data seem fairly normal, with maybe a slight leftward skew. It might be centered around \~15 deaths per 10,000 people with few to no outliers.

## A Simple Model

Let's start with a simpler model: Deaths as a function of Government Control

```{r}
deathgovmod<-glm.nb(Deaths~Gov_Control,data = covtotalslast)
autoplot(deathgovmod)
```
