---
title: "CRDT Covid"
format: html
editor: visual
---

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
ctcov<-read.csv(here("data","CRDT Data - CRDT.csv"))
glimpse(ctcov)
```

```{r}
ctcov<-ctcov|>
  rename_with(~str_remove(.,"thnicity_"))
ctcovlong <- ctcov |>
  pivot_longer(cols = matches("_"), # Keep column, reshape others
               names_to = "Race", # New column for race names
               values_to = "Count")# New column for counts
ctcovlong<-ctcovlong|>
  separate(Race, into = c("Type","Race"))
ctcov2<- ctcovlong |>
  pivot_wider(names_from = Type, values_from = Count)
```

```{r}
ctcov2 <- ctcov2 |>
  mutate(State = ifelse(State == "DC", "District of Columbia", 
                        state.name[match(State, state.abb)]))# Convert abbreviations to full names
ctcov2 <-ctcov2|>
  mutate(State = replace(State, is.na(State), "Territory"))
```

```{r}
ctcov2 <- ctcov2|>
  mutate(Region = case_when(
    State %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts", "Rhode Island", "Connecticut", "New York", "New Jersey", "Pennsylvania") ~ "Northeast",
    State == "District of Columbia" ~ "Midatlantic",  # Assign DC to a region
    State %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", "Iowa", "Minnesota", "Missouri", "North Dakota", "South Dakota", "Nebraska", "Kansas") ~ "Midwest",
    State %in% c("Delaware", "Maryland", "Virginia", "West Virginia", "North Carolina") ~ "Midatlantic",
    State %in% c("Kentucky", "South Carolina", "Tennessee", "Georgia", "Florida", "Alabama", "Mississippi", "Arkansas", "Louisiana", "Texas", "Oklahoma") ~ "South",
    State %in% c("Montana", "Idaho", "Wyoming", "Colorado", "New Mexico", "Arizona", "Utah", "Nevada") ~ "West",
    State %in% c("Washington", "Oregon", "California", "Alaska", "Hawaii") ~ "Pacific",
    State == "Territory" ~ "Territories of USA",
    TRUE ~ "Other"
  ))

```

```{r}
ctcov2 <- ctcov2|>
  mutate(Gov_Control = case_when(
    State %in% c("Alabama", "Alaska", "Arkansas", "Florida", "Georgia", 
                 "Idaho", "Indiana", "Iowa", "Mississippi", "Missouri", "Montana", 
                 "Nebraska", "North Dakota", "Ohio", "Oklahoma", "South Carolina", 
                 "South Dakota", "Tennessee", "Texas", "Utah", "West Virginia", 
                 "Wyoming", "New Hampshire") ~ "Red",
                 
    State %in% c("California", "Colorado", "Connecticut", "Delaware", "Hawaii", 
                 "Illinois", "Maine", "Massachusetts", "Nevada", "New Jersey", 
                 "New Mexico", "New York", "Oregon", "Rhode Island", "Vermont", 
                 "Virginia", "Washington", "District of Columbia") ~ "Blue",
                 
    State %in% c("Arizona", "Michigan", "Minnesota", "North Carolina", "Pennsylvania", 
                 "Wisconsin", "Kentucky", "Louisiana", "Maryland", "Kansas") ~ "Divided",
                 
    State %in% c("Territory") ~ "Territory",
                 
    TRUE ~ "Unknown"
  ))

```

```{r}
ctcov2$Date<-ymd(ctcov2$Date)
```

```{r}
library(tidycensus)

# Load 2020 Census state population data
state_pop_2020 <- get_decennial(
  geography = "state", 
  variables = "P1_001N",  # Total population from 2020 Census
  year = 2020, 
  sumfile = "pl"
)

# Rename columns for clarity
state_pop_2020 <- state_pop_2020 |>
  rename(State = NAME, Population = value)

# View the first few rows of the dataset
head(state_pop_2020)
ctcov2 <- ctcov2 %>%
  left_join(state_pop_2020, by = "State")  # Merging by State column

```

Munging Done!

## Now some ANOVAs

### Deaths vs Government Control

```{r}
ggplot(ctcov2,aes(Gov_Control,Deaths/Population))+
  geom_boxplot()
```

```{r}
#ggplot(ctcov3no,aes(Deaths))+
 # geom_histogram()
  #facet_wrap(~Gov_Control,ncol=1)
```

```{r}
mean_deaths <- mean(ctcov2$Deaths, na.rm = TRUE)
var_deaths <- var(ctcov2$Deaths, na.rm = TRUE)

print(paste("Mean:", mean_deaths, "Variance:", var_deaths))
```

```{r}
library(MASS)
ctcov2<-ctcov2|>filter(Deaths >= 0)
dgcmod <- glm.nb(Deaths/Population ~ Gov_Control, data = ctcov2)
#dgcmod<-lm(Deaths~Gov_Control,data = ctcov2)#,family = poisson)
autoplot(dgcmod)
```

Really bad model, let's fix outliers and just use the last date

```{r}
ctcov3<- ctcov2 |>
         filter(Date == "2021-03-07", Race != "Total",State!="Territory" )
outliers <- abs(ctcov3$Deaths) > 3 * sd(ctcov3$Deaths)
sum(outliers)
ctcov3no<- ctcov3[!outliers, ]
dgcmod<- glm.nb(Deaths/Population ~ Gov_Control, data = ctcov3no)
autoplot(dgcmod)
```

Better model but still bad, let's go with it for now

```{r}
anova(dgcmod)
summary(dgcmod)
```

There is not enough evidence to reject the null hypothesis, meaning that we cannot say government control affects the death rate by state. Perhaps when we figure out how to get better models, we will find one to fit better. Perhaps there was a bigger influence on the entire country at the time.

### Deaths vs Race

Plot it first

```{r}
ctcov3no2<-ctcov3no|>filter(Deaths/Cases<0.25&Race!="EHispanic"&Race!="ENonHispanic"&Race!="EUnknown")
ggplot(ctcov3no2, aes(Race,Deaths/Cases))+
  geom_boxplot()
```

```{r}
dcrmod<-glm(Deaths~Race, data = ctcov3no2,family=quasipoisson)
autoplot(dcrmod)
```

Good assumptions, but qq plot is weird

```{r}
anova(dcrmod)
summary(dcrmod)
```

Tukey

```{r}
library(emmeans)
library("multcomp")
library("multcompView")
tukey<-emmeans(dcrmod, pairwise ~ Race, adjust = "tukey")
emm <- tukey$emmeans
cldres <- as.data.frame(cld(emm))
print(cldres)
```

There is enough evidence to reject the null hypothesis that race has no effect on the deaths per cases ratio.

### Hospitalizations vs Gov Control

plot it

```{r}
ggplot(ctcov3no2,aes(Gov_Control,Hosp))+
  geom_boxplot()
```

model

```{r}
hgcmod<-glm(Hosp~Gov_Control,data =ctcov3no2,family=poisson)
autoplot(hgcmod)
```

Worry about q-q plot later

```{r}
anova(hgcmod)
summary(hgcmod)
```

```{r}
tukey2<-emmeans(hgcmod, pairwise ~ Gov_Control, adjust = "tukey")
emm2 <- tukey2$emmeans
cldres2 <- as.data.frame(cld(emm2))
print(cldres2)
```

There is enough evidence to reject the null hypothesis that gov control does not affect hospitalizations.

## Now some ANCOVAs

Deaths\~Govt control and Cases two way anova

```{r}
ctcov3<- ctcov2 |>
         filter(Date == "2021-03-07", Race != "Total",State!="Territory" )
```

```{r}
ggplot(ctcov3, #|>
        # filter(Cases >= quantile(Cases, 0.25,na.rm = TRUE) - 1.5 * IQR(Cases,na.rm = TRUE) & 
         #       Cases <= quantile(Cases, 0.75,na.rm = TRUE) + 1.5 * IQR(Cases,na.rm = TRUE)),
       aes(Cases,Deaths, colour = Gov_Control)) +
  scale_color_manual(values = c(Blue='blue',Red='red',Divided='yellow'))+
  geom_point()
```

What are those weird points on the left and right?

```{r}
filter(ctcov3, Deaths>15000&Cases<250000)
```

```{r}
filter(ctcov3, Deaths<5000&Cases>2500000)
```

Should point out that outliers at low cases/high deaths are red states

```{r}
dcgmod<-lm(Deaths~Cases*Gov_Control,data = ctcov3)
autoplot(dcgmod)
```

# New csv

```{r}
#write.csv(ctcov2, file = "first_merged_CRDT.csv")
```
